---
title: "1974 Motor Trends Analysis"
author: "Isioma Mabatah"
date: "9/2/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
---

# Overview

This report will be describing vehicles listed in the **1974 Motor Trends US Magazine**. The assignment of interest comes from [RPubs.com](https://rpubs.com/neros/61800). The objective is to explore the relationship between **mpg** and the other features listed in the dataset.

Particular Questions of Interest

1. Is an automatic or manual transmission better for **mpg**.
2. Quantify the **mpg** difference between automatic and manual transmissions.

## Libraries

```{r loading packages, message=FALSE}
library(dplyr)
library(ggplot2)
library(factoextra)
library(FactoMineR)
library(plotly)
library(broom)
```


# Analysis

## Data

The `mtcars` dataset is already provided in R. The Documentation can be found [here](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars)

```{r loading_data}
set.seed(1993)
df <- mtcars
df %>% glimpse()
```


### Description of Features

```{r description}
df_description <- tibble(
  Feature = c("mpg","cyl","disp","hp","drat","wt","qsec","vs","am","gear","carb"),
  Description = c("Miles / US Gallon",
                  "Number of Cylinders (4,6,8)",
                  "Displacement - Measures the overall volume in the engine as a factor of cylinder circumfrance, depth and total number of cylinders. 'Proxy for the total amount of power the engine can generate.'",
                  "Horsepower",
                  "Rear Axle Ratio - The number of turns of the drive shaft for every one rotation of the wheel axle",
                  "Weight (lb/1000)",
                  "1/4 time - A performance measure, primarily of acceleration",
                  "Engine Cylinder Configuration - (VShape = 0 | StraightLine = 1)",
                  "Transmission Type - (Auto = 0 | Manual = 1)",
                  "Number of Forward Gears - Auto = (3 | 4), Manual = (4 | 5)",
                  "Number of Carburetors - Engines with higher displacement typically have higher barrel configuration")
) %>% knitr::kable()

df_description
```


Refer to [RPubs](https://rpubs.com/neros/61800) for a more in depth description

### Preprocessing


```{r glimpse}
df <- df %>% mutate(
  cyl = as.factor(cyl),
  vs = as.factor(ifelse(vs == 1, "S","V")),
  am = as.factor(ifelse(am == 1, "M","A")),
  gear = as.factor(gear),
  name = rownames(df)
)
# Split to Train and Testset 
set <- sample(1:2, replace = TRUE,
              size = nrow(df),
              prob = c(0.6,0.4))
df_train <- df[set == 1,]
df_test <- df[set == 2,]
df_train %>% skimr::skim()
```

Note: 

  - Data split (60/40)
  
  - cyl, vs, am, and gear were changed to factors


## Exploratory Data Analysis

### PCA

```{r pca_object, fig.align='center'}
# Construct PCA Object: 
df_train_PCA <- 
  df_train %>% select(-cyl,-vs,-am,-gear) %>% 
  PCA(graph = FALSE,
      # Target (MPG)
      quanti.sup = 1,
      # Vehicle Names
      quali.sup = 8)
```


```{r Mpg_Variables, fig.align='center', fig.cap="Figure 1"}
df_train_PCA$quanti.sup$cor[,1:2] %>% round(2); df_train_PCA$quanti.sup$cos2[,1:2] %>% round(2)

df_train_PCA %>% 
  fviz_pca_var(col.quanti.sup = "navyblue", invisible = "var") + 
  theme_minimal() +
  labs(
    title = "MPG Summary",
    x = "MPG Cor: -0.877 | Quality of Representaion: 76%",
    y = "MPG Cor:  0.254 | Quality of Representaion: 6%"
  ) +
  theme(
    plot.background = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )
```


```{r Eigenvalues, fig.align='center',fig.cap="Figure 2"}
round(df_train_PCA$eig,2)
df_train_PCA %>% 
  fviz_eig(addlabels = TRUE, ylim = c(0,70)) +
  labs(
    title = "Principal Components 1 and 2 explains about 86% of the variance in the data",
    x = "Principal Components",
    y = element_blank()
  ) +
  theme_minimal() +
  theme(
    plot.background = element_blank(),
    axis.text.y = element_blank()
  )
```

#### Principal Components: Total Contribution

```{r PC1, fig.align='center', fig.cap="Figure 3.1"}
df_train_PCA$var$contrib[,1] %>% sort(decreasing = TRUE) %>% round(2)
df_train_PCA %>% 
  fviz_contrib(choice = "var", axes = 1) + 
  labs(
    title = "",
    x = element_blank(),
    y = element_blank()
  ) +
  theme_minimal() +
  theme(
    plot.background = element_blank(),
    axis.text.y = element_blank()
  )
```

**Principal Component 1:**

 - These variables seems to correlate with the overall power of a vehicle. I would assume a vehicle with high horsepower has a high displacement. Weight would increase as well.
  

```{r PC2, fig.align='center', fig.cap="Figure 3.2"}
df_train_PCA$var$contrib[,2] %>% sort(decreasing = TRUE) %>% round(2)
df_train_PCA %>% 
  fviz_contrib(choice = "var", axes = 2) +
  labs(
    title = "",
    x = element_blank(),
    y = element_blank()
  ) +
  theme_minimal() +
  theme(
    plot.background = element_blank(),
    axis.text.y = element_blank()
  )
```


**Principal Component 2:**

  - drat "Rear axle ratio":  as this ratio goes up, towing capacity increases which in turn reduces the fuel economy of the vehicle
  
  - qsec "Measure of Acceleration": [edmunds.com](https://www.edmunds.com/car-buying/how-to-choose-the-right-axle-ratio-for-your-pickup-truck.html#:~:text=The%20term%20refers%20to%20the,each%20turn%20of%20a%20wheel.). 
  
  - carb "Carburetor": used to mix air with gasoline by using an optimal amount of fire to burn the gas. A good Carburator will be able to move fuel thoughout the vehicle efficiently thus increasing power to improve speed [CARBURETORS | How They Work](https://www.youtube.com/watch?v=1ohL6KgasjA)
  
  _All in all based on these variable descriptions I would say that this PC is more associated with how a efficient a vehicle uses its fuel in terms of performance_ 


#### PCA-Categories

```{r PCA_Cylinder, fig.align='center', fig.cap="Figure 4.1"}
df_train_PCA %>% fviz_pca_biplot(
  geom = "point", pointshape = 21, fill.ind = "gray", col.ind = df_train$cyl,
  col.var = "black", alpha.var = "cos2", col.quanti.sup = "navyblue",
  addEllipses = TRUE, ellipse.type = "confidence",
  palette = c("lightsteelblue4","lightskyblue","lightseagreen"),
  legend.title = "", mean.point = FALSE
) +
  labs(
    title = "Cylinder",
    x = element_blank(),
    y = element_blank(),
    alpha = "Quality of Representation"
  ) +
  theme(
    plot.background = element_blank(),
    legend.position = "top"
  )
```

_Note:_

 _- While the cylinder of a vehicle increases so does the Horsepower, Displacement and Weight, which in turn leads to a lower mpg_


```{r PCA_VS, fig.align='center', fig.cap="Figure 4.2"}
df_train_PCA %>% fviz_pca_biplot(
  geom = "point", pointshape = 21, fill.ind = "gray", col.ind = df_train$vs,
  col.var = "black", alpha.var = "cos2", col.quanti.sup = "navyblue",
  addEllipses = TRUE, ellipse.type = "confidence",
  palette = c("lightsteelblue4", "lightseagreen"),
  legend.title = "", mean.point = FALSE
) +
  labs(
    title = "Engine Configuration",
    x = element_blank(),
    y = element_blank(),
    alpha = "Quality of Representation"
  ) +
  theme(
    plot.background = element_blank(),
    legend.position = "top"
  )
```

_Note:_

 _- Straightline engines seem to have a lower Horsepower, Displacement, and Weight compared to V-Shaped engines, which leads to a higher mpg_


```{r PCA_Gear, fig.align='center', fig.cap="Figure 4.3"}
df_train_PCA %>% fviz_pca_biplot(
  geom = "point", pointshape = 21, fill.ind = "gray", col.ind = df_train$gear,
  col.var = "black", alpha.var = "cos2", col.quanti.sup = "navyblue",
  addEllipses = TRUE, ellipse.type = "confidence",
  palette = c("lightsteelblue4","lightskyblue","lightseagreen"),
  legend.title = "", mean.point = FALSE
) +
  labs(
    title = "Gear",
    x = element_blank(),
    y = element_blank(),
    alpha = "Quality of Representation"
  ) +
  theme(
    plot.background = element_blank(),
    legend.position = "top"
  )
```

_Note:_

 _- A gear 4 vehicle is associated with a better mpg on avg than a gear 3 and 5 vehicles_
 
 _- A gear 5 vehicles with higher drat "Rear Axle Ratio" leads to a higher mpg._


```{r PCA_AM, fig.align='center', fig.cap="Figure: 4.4"}
df_train_PCA %>% fviz_pca_biplot(
  geom = "point", pointshape = 21, fill.ind = "gray", col.ind = df_train$am,
  col.var = "black", alpha.var = "cos2", col.quanti.sup = "navyblue",
  addEllipses = TRUE, ellipse.type = "confidence",
  palette = c("tomato", "seagreen"),
  legend.title = "", mean.point = FALSE
) +
  labs(
    title = "Transmission",
    subtitle = "I used a different color scheme to highlight that this is the particular catergory of interest",
    x = element_blank(),
    y = element_blank(),
    alpha = "Quality of Representation"
  ) +
  theme(
    plot.background = element_blank(),
    legend.position = "top"
  )
```


_Note:_

 _- Manual Transmission is a associated with a better mpg than Automatic._


#### PCA - Biplot

```{r Biplot, fig.align='center', fig.cap="Figure: 5"}
df_train_PCA %>% 
  fviz_pca_biplot(invisible = "ind", col.var = "black", alpha.var = "cos2",
                  col.quanti.sup = "navyblue") %>% 
  fviz_add(df_train_PCA$quali.sup$coord, repel = TRUE,
           geom = "text", color = "lightsteelblue4") +
  labs(
    title = "Vehicle Names",
    x = element_blank(),
    y = element_blank(),
    alpha = "Quality of Represention"
  ) +
  theme(
    plot.background = element_blank(),
    legend.position = "top"
  )
```

_Note:_

 _- Honda Civic "what I drive currently :)" is a lower weighted vehicle with good mpg compared to the others in the dataset_
 
 _- Cadillac Fleetwood is a higher weighted vehicle which is associated with a lower mpg compared to the others in the data_
 
 _- Maserati Bora is a high carb vehicle associated with a worse mpg compared to the others in the data_
 
 _- Merc 230 has a good "Measure of Acceleration" compared to the other vehicles in the data_
 

### Plots

```{r plot_cor, fig.align='center', warning=FALSE, fig.cap="Figure: 6.1"}
df_train %>% select(mpg,disp,hp,wt,qsec,drat,carb) %>% 
  cor() %>% round(2)
df_train %>% 
  plot_ly(color = ~am, 
          hoverinfo = "text",
          text = ~paste("Name:", name, "<br>", am),
          colors = c("tomato", "seagreen")) %>% 
  add_trace(
    type = "splom",
    dimensions = list(
      list(label = "mpg", values = ~mpg),
      list(label = "wt", values = ~wt),
      list(label = "drat", values = ~drat)
    )
  ) %>%
  style(showlowerhalf = FALSE, showlegend = FALSE) %>% 
  layout(title = "Correlation Plot")
```

Findings: mpg by am (transmission)

 - controlling for wt, it seems as though the mpg of a vehicle varies evenly no matter the type of transmission. However their seem to be a few outliers for Automatic vehicles at the high end of wt.
 
 - controlling for drat the data also varies. It looks like their might be one potential outlier at the high end of drat.
 
 - _using the lasso select tool_ if you highlight the potential outliers at wt(5,5), you can see that those vehicles is associated with a low drat "Rear Axle Ratio" and mpg


_Note:_

 _- If you look at Figure 3.1: PC1 was best represented by disp, hp, and wt. I choose **wt** because it is, even though not by very much, the most correlated with mpg_
 
 _- If you look at Figure 3.2: PC2 was best represented by qsec, drat, and carb. I choose **drat** for similar reasons_


```{r plot, fig.align='center', warning=FALSE, fig.cap = "Figure: 6.2"}
df_train %>% 
  plot_ly(x = ~wt, y = ~drat, z = ~mpg,
          hoverinfo = "text",
          text = ~paste("Mpg:", mpg, "<br>",
                        "Weight:", wt, "<br>",
                        "Drat:", drat, "<br>",
                        "Transmission:", am, "<br>",
                        "Name:", name),
          colors = c("tomato", "seagreen")
          ) %>% 
  add_markers(color = ~am, showlegend = FALSE) %>% 
  layout(xaxis = list(title = "Weight (lb/1000)"),
         yaxis = list(title = "Miles per US Gallon"),
         title = "Automatic vs Manual",
         plot_bgcolor = toRGB("gray90"))
```


Findings: mpg ~ am + drat + wt (transmission)

 - controling for drat and wt, you can see more of a relationship between Transmission Types.
 
 - Manual: tend to have lower weight but high drat which lends to a better mpg compared to auto
 
 - Automatic: tend to have high wt and lower drat which lead to a worse mpg compared to manual
 
 _This graph is also interactive to feel free to play around with it_
 

# Modeling 

To Quantify the difference I will be using a regression model

### Model Fit

```{r assessment}
mod <- lm(mpg ~ am + drat + wt, data = df_train)
mod %>% tidy()
```

_Note:_

 _- The intercept represents Automatic_

 _- Controlling for other the variables in the model, only wt seems to be significant_

```{r interaction}
mod_2 <- lm(mpg ~ am + drat + am:wt, data = df_test)
mod_2 %>% tidy()
```

_Note:_

 _- all terms are insignificant at the **0.05** level_
 

```{r model_3}
mod_3 <- lm(mpg ~ am + am:wt, data = df_train)
mod_3 %>% tidy()
```

_Note:_
 
 _- controlling for all the the variables in the model, when the transmission is **Manual** a 1 unit increase in weight leads to a decrease in mpg at a higher rate compared to Automatic._

 _- It looks like this is a good model at the **0.05 and 0.01 significance level**, but I'll stick to the 0.05 level_
 
```{r}
mod_3 %>% glance() %>% select(r.squared, rmse = sigma)
```
 

_Note:_ 

 _- At the **0.05 significance level** this model explains **`r glance(mod_3) %>% select(r.squared) %>% round(2)`** of the variance._

 _- Training RMSE: **`r glance(mod_3) %>% select(rmse = sigma) %>% round(2)`**_

### Diagnostics

```{r normality, fig.align='center', fig.cap="Figure: 7.1"}
mod_3 %>% augment() %>% cbind(name = df_train$name) %>% 
  ggplot(aes(.fitted, .resid, color = am)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "black") +
  geom_text(aes(label = name)) +
  scale_color_manual(values = c("tomato", "seagreen")) +
  labs(
    title = "Apart from a potential outlier, Variance seems to be Homoscedastic",
    x = "Fitted Values",
    y = "Residuals",
    color = "Transmission"
  ) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r Nomality Standardized, fig.align='center', fig.cap="Figure: 7.2"}
mod_3 %>% augment() %>% cbind(name = df_train$name) %>% 
  ggplot(aes(.fitted, .std.resid, color = am)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "black") +
  geom_text(aes(label = name)) +
  scale_color_manual(values = c("tomato", "seagreen")) +
  xlim(c(10,32)) +
  labs(
    title = "No Outliers - Merc 240D is still with 3 standard deviations",
    x = "Fitted Values",
    y = "Standardized Residuals",
    color = "Transmission"
  ) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```


```{r linearity, fig.align='center', fig.cap="Figure 7.3"}
mod_3 %>% augment() %>% cbind(name = df_train$name) %>% 
  ggplot(aes(.fitted, mpg, color = am)) +
  geom_point() +
  geom_text(aes(label = name)) +
  geom_abline(intercept = 0, slope = 1, color = "black") +
  scale_color_manual(values = c("tomato", "seagreen")) +
  xlab(c(5,40)) +
  labs(
    title = "Normality Check",
    x = "Fitted Values",
    y = "Actual Values",
    color = "Transmission"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## Predictions

```{r predictions}
mod_3_pred <- mod_3 %>% predict(df_test) %>% round(2)
df_test %>% 
  select(name, am, wt, mpg) %>% 
  cbind(mpg_pred = mod_3_pred) %>% 
  knitr::kable()

Metrics::rmse(actual = df_test$mpg,
              predicted = mod_3_pred) %>% 
  round(2)
```

_Note:_

 _- Test RMSE: **`r Metrics::rmse(actual = df_test$mpg,predicted = mod_3_pred) %>% round(2)`**_

# Results

## 1. Is an automatic or manual transmission better for **mpg**.

 Based on the sample taken from the **1974 Motor Trends Analysis**. 
 
  - If your vehicle weighs below 2,500lb (2.5 * 1000). I would say go for the manual transmission because it has a better mpg. 
  
  - If the weight of a vehicle is above 2.500lbs there not much of a difference between manual and automatic

## 2. Quantify the **mpg** difference between automatic and manual transmissions.

```{r}
mod_3
```

 - Rsquared: `r glance(mod_3) %>% select(r.squared) %>% round(2)` | Rmse: `r Metrics::rmse(actual = df_test$mpg,predicted = mod_3_pred) %>% round(2)`
 
 - According to the model Automatic Transmission vehicles have an avg mpg of `r mod_3$coefficients[1] %>% round(2)`. As weight increases by 1 unit _(1000/lbs)_, for an Automatic vehicle the mpg will drop by `r mod_3$coefficients[3] %>% round(2)`
 
 - If the vehicle has a Manual Transmission the mpg increases by `r mod_3$coefficients[2] %>% round(2)`. As weight increases by 1 unit _(1000/lbs)_, for a Manual vehicle the mpg will drop by `r mod_3$coefficients[4] %>% round(2)`
 

_Note: The sample size of the `mtcars` dataset was very small so I would be cautious to generalize these findings to the entire population_ 











